---
title: "R Notebook"
output: html_notebook
---


```{r}
#BiocManager::install("DESeq2")
#BiocManager::install("airway")

library(DESeq2)
library(airway)
library(dplyr)
data(airway)
airway
head(assay(airway))

```


```{r}
assay(airway,"counts")[1:3,1:3]
airway$dex
airway$dex<- relevel(airway$dex,"untrt")
airway$dex
colData(airway)

```

There is rich information about which gene model was used for each gene:
```{r}
granges(airway)
```

1. Count Normalization
```{r}
# Create DESeq2Dataset object
dds <- DESeqDataSet(airway, design = ~ dex)
##############
dds2<- estimateSizeFactors(dds)
sizeFactors(dds2)
dds2<- counts(dds2, normalized =T)
head(dds2)
```


2. Exploratory data analysis (PCA & heirarchical clustering) - identifying outliers and sources of variation in the data:
```{r}
# Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)

# Plot PCA 
plotPCA(rld, intgroup="dex")

# Extract the rlog matrix from the object
rld_mat <- assay(rld)

# Compute pairwise correlation values
rld_cor <- cor(rld_mat)


# Plot heatmap
install.packages("pheatmap", "RColorBrewer", "viridis")
library(pheatmap)
pheatmap(rld_cor)
```

3. Run DESeq2
```{r}
#Fitting the model
dds<- DESeq(dds)
estimateSizeFactors(dds)

#Check the fit of the dispersion estimates:
plotDispEsts(dds)
```


5. Create contrasts to perform Wald testing on the shrunken log2 foldchanges between specific conditions:
```{r}
# Output results of Wald test for contrast
res <- results(dds,contrast = c("dex","trt","untrt"), alpha=0.05)
res <- res[order(res$padj),]
head(res)
```


6. Output significant results:
```{r}
# Turn the results object into a data frame
res <- results(dds)

head(res)
res_df <- data.frame(res)
head(res_df)
summary(res)
```


7. Visualization
```{r}
plotMA(res, ylim=c(-4,4), na.rm=T)

#Annotation
library("AnnotationDbi")
library("org.Hs.eg.db")
columns(org.Hs.eg.db)
res$symbol <- mapIds(org.Hs.eg.db, 
                     keys=row.names(res), 
                     column="SYMBOL", 
                     keytype="ENSEMBL",
                     multiVals="first")
res$entrez <- mapIds(org.Hs.eg.db, 
                     keys=row.names(res), 
                     column="ENTREZID", 
                     keytype="ENSEMBL",
                     multiVals="first")
resOrdered <- res[order(res$padj),]
head(resOrdered)

```




